---
title: "Frequently Asked Questions"
output: 
  html_document:
    number_sections: true
    toc: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Installation

- [How do I install an R package for the first time?](https://drieslab.github.io/Giotto/articles/installation.html)

- [Can I install Python components manually?](https://drieslab.github.io/Giotto/articles/configuration.html)

# Data Availability

- Where can I find seqFISH+ and other ready-to-use datasets?

Checkout our [GiottoData](https://github.com/drieslab/GiottoData) extension package to find already preprocessed datasets and Giotto mini Objects.

-  Where else can I find more spatial datasets?

Checkout the following for more spatial-omics data:

 - [Aquila](https://aquila.cheunglab.org/view)
 - Tencent\'s [SODB](https://gene.ai.tencent.com/SpatialOmics/)
 - [PySODB](https://pysodb.readthedocs.io/en/latest/), a python interface for the SODB

- How can I automatically download tutorial datasets?

Use `getSpatialDataset()` from GiottoData:

```{r, eval=FALSE}
# Ensure Giotto Suite is installed
if(!"Giotto" %in% installed.packages()) {
  pak::pkg_install("drieslab/Giotto")
}

library(Giotto)

# Ensure Giotto Data is installed
if(!"GiottoData" %in% installed.packages()) {
  pak::pkg_install("drieslab/GiottoData")
}

library(GiottoData)


# choose your directory
my_working_dir = getwd()

# merFISH example:

# standard download data to working directory
getSpatialDataset(dataset = 'merfish_preoptic', 
                  directory = my_working_dir)

# use wget to  download data to working directory (much faster)
getSpatialDataset(dataset = 'merfish_preoptic', 
                  directory = my_working_dir, 
                  method = 'wget')

# avoid certification issues with wget
getSpatialDataset(dataset = 'merfish_preoptic', 
                  directory = my_working_dir, 
                  method = 'wget', 
                  extra = '--no-check-certificate')
```


# Matrix error

- Why am I getting the error `function 'as_cholmod_sparse' not provided by package 'Matrix'`?

Matrix underwent a breaking change after versoin 1.6-2. Reinstalling _irlba_ from source will fix this.
See this [issue](https://github.com/bwlewis/irlba/issues/70) for more info.

```{r, eval=FALSE}
install.packages("irlba", type = "source")
```

# Arrow and ZSTD

- Why am I getting an error about arrow and ZSTD support when loading Xenium or VisiumHD data?

*arrow* is a suggested package Giotto uses to open `parquet` files. The parquet files that 10X provides use ZSTD compression which the default *arrow* installation may not provide.
The following code will check for whether arrow is installed with ZSTD support, and reinstall if it is not present.

```{r, eval=FALSE}
has_arrow <- requireNamespace("arrow", quietly = TRUE)
zstd <- TRUE
if (has_arrow) {
    zstd <- arrow::arrow_info()$capabilities[["zstd"]]
}
if (!has_arrow || !zstd) {
    Sys.setenv(ARROW_WITH_ZSTD = "ON") 
    install.packages("assertthat", "bit64")
    install.packages("arrow", repos = c("https://apache.r-universe.dev"))
}
```


